---
title: "Setting up requirements and packages"
author: "Al Xin"
date: "2025-09-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Loading from `renv`

We manage the packages and versions in this project with `renv`. Currently, `renv` is configured to work with R version 4.4.2. Some of the included packages were built with later versions of R, but they should still work within the project. To load all the dependencies, run the following:

```{r}
if (!require("renv")) install.packages("renv")
renv::restore()
```

More information about `renv` can be found in its [documentation](https://rstudio.github.io/renv/articles/renv.html).

## Downloading from source

If `renv::restore()` fails to install `fastFMM`, it may be necessary to download them separately. We built these figures with the development version of `fastFMM`, available on GitHub at the repository [awqx/fastFMM](https://github.com/awqx/fastFMM). 

```{r}
if (!require("devtools")) install.packages("devtools")
suppressMessages(
  devtools::install_github("awqx/fastFMM")
)
```



