---
title: 'Figure 3: Variable trial lengths in D2+ PVT mice experiments'
author: "Al Xin"
date: "2025-08-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library(dplyr))
```

## Overview

## Data processing

Data processing can be found in `data/3-d2pvt.R`. Note that the raw data is currently not published publicly, so the code will not work out of the box. 

```{r}
dat <- read.csv("data/3-d2pvt.csv") %>%
  mutate(outcome = as.factor(outcome)) %>%
  mutate(outcome = relevel(outcome, ref = "H-H2O"))
```

### Model-fitting

The model formula is informed by previous analysis. 

```{r}
hsms_hh2o_rwd <- fastFMM::fui(
  photometry ~ outcome * reward + (outcome | id), 
  dat, 
  concurrent = T
)
```

### Plots

```{r, eval = F}
png(filename = paste0(img_dir, "hsms_hh2o_rwd.png"), width = 600, height = 400)
fastFMM::plot_fui(hsms_hh2o_rwd)
dev.off()
```

For more flexibility, the plots can also be generated in `ggplot2`. 

```{r}
source("fun/plottable_fmm.R")
# mutate s based on zero var col offset

cov_names <- c(
  "Baseline approach", 
  "SMS approach", 
  "Baseline reward", 
  "SMS reward − baseline reward"
)

hsms_hh2o_df <- plottable_fmm(
  hsms_hh2o_rwd, 
  cov_names = cov_names
) %>%
  mutate(s = s + 30)

# Get the correct s offset

# Get times from the original CSV
hsms_hh2o_orig <- read.csv("data/d2pvt/D2PVT_FLMM_input_HSMSvsHH2O.csv")
times <- as.numeric(
  gsub("^X(\\.)*", "", colnames(hsms_hh2o_orig)[7:ncol(hsms_hh2o_orig)])
)
# Convert to negative
times[1:24] <- -times[1:24]
times_df <- data.frame(s = 1:length(times), time = times)

hsms_hh2o_df <- left_join(hsms_hh2o_df, times_df, by = "s") %>%
  mutate(cov = factor(cov, cov_names)) 

hsms_hh2o_labs <- data.frame(
  # label = c("A", "B", "C", "D"), 
  label = c("i", "ii", "iii", "iv"), 
  cov = cov_names
  # x = 1.5, 
  # y = c(3.5, 5, 2, 4.5)
) %>%
  mutate(cov = factor(cov, cov_names))

ggplot(hsms_hh2o_df, aes(x = time, y = val)) + 
  facet_wrap(vars(cov), nrow = 1, scales = "free_y") + 
  geom_line() + 
  geom_ribbon(aes(ymin = val - ci, ymax = val + ci), alpha = 0.2) + 
  geom_ribbon(aes(ymin = val - joint_ci, ymax = val + joint_ci), alpha = 0.2) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  labs(
    x = "Trial time (s)", 
    y = "Fixed coefficient"
  ) + 
  theme(
    panel.border = element_blank(), 
    panel.grid.minor = element_blank()
  )

ggsave(paste0(img_dir, "hsms_hh2o-A.png"), width = 10, height = 2.5)
```

## Non-concurrent model with latency


Instead of `reward`, the non-concurrent equivalent is `latency`. 

```{r}
hsms_hh2o_lat <- fastFMM::fui(
  photometry ~ outcome * latency + (outcome | id), 
  dat, 
  concurrent = F
)
```

### Plots

```{r}
png(filename = paste0(img_dir, "hsms_hh2o_lat.png"), width = 600, height = 400)
fastFMM::plot_fui(hsms_hh2o_lat)
dev.off()
```

And again in `ggplot2`:

```{r ggplot lat}
source("fun/plottable_fmm.R")
# mutate s based on zero var col offset

cov_lat <- c("Baseline", "SMS", "Latency", "SMS * latency")

hsms_hh2o_lat_df <- plottable_fmm(
  hsms_hh2o_lat, 
  cov_names = cov_lat
) # %>%
#   mutate(s = s + 30)

hsms_hh2o_lat_df <- left_join(hsms_hh2o_lat_df, times_df, by = "s") %>%
  mutate(cov = factor(cov, cov_lat))

hsms_hh2o_lat_labs <- data.frame(
  label = c("i", "ii", "iii", "iv"),
  cov = cov_lat
) %>%
  mutate(cov = factor(cov, cov_lat))

ggplot(hsms_hh2o_lat_df, aes(x = time, y = val)) + 
  facet_wrap(vars(cov), nrow = 1, scales = "free_y") + 
  geom_line() + 
  geom_ribbon(aes(ymin = val - ci, ymax = val + ci), alpha = 0.2) + 
  geom_ribbon(aes(ymin = val - joint_ci, ymax = val + joint_ci), alpha = 0.2) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  labs(
    x = "Trial time (s)", 
    y = "Fixed coefficient"
  ) + 
  theme(
    panel.border = element_blank(), 
    panel.grid.minor = element_blank()
  )

ggsave(paste0(img_dir, "hsms_hh2o-B.png"), width = 10, height = 2.5)
```

## Linear mixed model with average and sum of signal

Instead of a functional linear mixed model, we can fit a linear mixed model. The functional photometry measure is replaced by a scalar of either the average photometry signal in the reward period or the sum.

### Reprocess data

```{r}
dat <- read.csv(paste0(proc_dir, "hsms_hh2o.csv")) %>%
  mutate(outcome = as.factor(outcome)) %>%
  mutate(outcome = relevel(outcome, ref = "H-H2O")) %>%
  mutate(
    session = session - 1, 
    trial = trial - 1, 
    latency = latency - mean(latency)
  ) %>%
  mutate(
    trial = trial / max(trial), 
    session = session / max(session)
  )
```

The inverse of the `approaching` functional covariate (i.e., `reward`) indicates which photometry signals are relevant. Here, the average of photometry signals in the reward period (`photo_avg`) and the sum in the reward period (`photo_sum`) are reported.

```{r}
reward_df <- dat[, grep("^approaching", colnames(dat))] %>%
  as.matrix() 
reward_df <- abs(reward_df - 1) 

photometry_df <- dat[, grep("^photometry", colnames(dat))] %>%
  as.matrix()

# photo average
photo_incl <- photometry_df * reward_df
dat$photo_avg <- rowSums(photo_incl) / rowSums(reward_df)

# sum of signals over reward
dat$photo_sum <- rowSums(photo_incl)
dat$photo_sum <- dat$photo_sum / sd(dat$photo_sum)
```

### Building LMMs with `lme4::lmer`

```{r}
lmm_avg <- lme4::lmer(
  formula = photo_avg ~ outcome * latency + (outcome | id),
  data = dat,
  control = lme4::lmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 5000),
    check.rankX = "ignore"
  )
)

lmm_sum <- lme4::lmer(
  formula = photo_sum ~ outcome * latency + (outcome | id),
  data = dat,
  control = lme4::lmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 5000),
    check.rankX = "ignore"
  )
)
```

Dot-whisker plots can be created with `ggplot2`, where "whiskers" are CIs and "boxes" (thicker whiskers) are S.E.

#### Average signal

```{r}
lmm_avg_outs <- summary(lmm_avg)
coef_nm <- c(
  "Intercept", 
  "SMS", 
  "Latency", 
  "SMS * latency"
)

lmm_avg_ci <- suppressMessages(confint(lmm_avg))

lmm_avg_df <- data.frame(
  coef_nm = coef_nm, 
  fixfx = lmm_avg_outs$coefficients[, 1], 
  sd = lmm_avg_outs$coefficients[, 2], 
  ci_lo = lmm_avg_ci[5:8, 1], 
  ci_hi = lmm_avg_ci[5:8, 2]
) %>%
  mutate(coef_nm = factor(coef_nm, rev(coef_nm)))

ggplot(lmm_avg_df, aes(x = coef_nm, y = fixfx)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi), width = 0) + 
  geom_errorbar(
    aes(ymin = fixfx - sd, ymax = fixfx + sd), 
    width = 0,
    size = 1
  ) + 
  theme(
    panel.border = element_blank(), 
    panel.grid.minor = element_blank()
  ) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  labs(
    y = "Fixed effect coefficient", 
    x = "", 
    title = "Reward signal average"
  ) + 
  coord_flip()
ggsave(paste0(img_dir, "lmm_avg.png"), width = 5, height = 2.5)
```

#### Sum of signal
 
```{r}
lmm_sum_outs <- summary(lmm_sum)
coef_nm <- c(
  "Intercept", 
  "SMS", 
  "Latency", 
  "SMS * latency"
)

lmm_sum_ci <- suppressMessages(confint(lmm_sum))

lmm_sum_df <- data.frame(
  coef_nm = coef_nm, 
  fixfx = lmm_sum_outs$coefficients[, 1], 
  sd = lmm_sum_outs$coefficients[, 2], 
  ci_lo = lmm_sum_ci[5:8, 1], 
  ci_hi = lmm_sum_ci[5:8, 2]
) %>%
  mutate(coef_nm = factor(coef_nm, rev(coef_nm)))

ggplot(lmm_sum_df, aes(x = coef_nm, y = fixfx)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi), width = 0) + 
  geom_errorbar(
    aes(ymin = fixfx - sd, ymax = fixfx + sd), 
    width = 0,
    size = 1
  ) + 
  theme(
    panel.border = element_blank(), 
    panel.grid.minor = element_blank()
  ) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  labs(
    y = "Fixed effect coefficient", 
    x = "", 
    title = "Reward signal sum"
  ) + 
  coord_flip()

ggsave(paste0(img_dir, "lmm_sum.png"), width = 5, height = 2.5)
```
 

```{r}
ggplot(dat, aes(x = latency, y = photo_avg)) + 
  geom_point(alpha = 0.2) + 
  geom_smooth(method = "lm", se = T) + 
  facet_wrap(vars(outcome)) + 
  # ylim(c(-10, 10)) + 
  coord_cartesian(ylim = c(-10, 10))

# ggplot(dat, aes(x = latency, y = photo_max)) + 
#   geom_point(alpha = 0.2) + 
#   geom_smooth(method = "lm", se = T)
```

### Building LMMs with `nlme::lme`

The above models with `lme4::lmer` do not provide p-values. We can recompute models with `nlme::lme`. 

#### Models

```{r}
lme_avg <- nlme::lme(
  fixed = photo_avg ~ outcome * latency, 
  random = ~ outcome | id, 
  data = dat
)

lme_sum <- nlme::lme(
  fixed = photo_sum ~ outcome * latency, 
  random = ~ outcome | id, 
  data = dat
)
```

#### Plots

Point estimates, standard deviation, and confidence intervals are placed in a plottable data frame. 

```{r}
coef_nm <- c(
  "Intercept", 
  "SMS", 
  "Latency", 
  "SMS * latency"
)

lme_sum_outs <- summary(lme_sum)
lme_sum_ci <- data.frame(nlme::intervals(lme_sum)$fixed)
lme_sum_df <- data.frame(
  coef_nm = coef_nm, 
  fixfx = lme_sum_outs$coefficients$fixed, 
  sd = lme_sum_outs$varFix %>% diag() %>% sqrt(), 
  ci_lo = lme_sum_ci$lower, 
  ci_hi = lme_sum_ci$upper
) %>%
  mutate(coef_nm = factor(coef_nm, rev(coef_nm)))

lme_avg_outs <- summary(lme_avg)
lme_avg_ci <- data.frame(nlme::intervals(lme_avg)$fixed)
lme_avg_df <- data.frame(
  coef_nm = coef_nm, 
  fixfx = lme_avg_outs$coefficients$fixed, 
  sd = lme_avg_outs$varFix %>% diag() %>% sqrt(), 
  ci_lo = lme_avg_ci$lower, 
  ci_hi = lme_avg_ci$upper
) %>%
  mutate(coef_nm = factor(coef_nm, rev(coef_nm)))
```

Box and whiskers for photometry average: 

```{r}
ggplot(lme_sum_df, aes(x = coef_nm, y = fixfx)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi), width = 0) + 
  geom_errorbar(
    aes(ymin = fixfx - sd, ymax = fixfx + sd), 
    width = 0,
    linewidth = 1.2
  ) + 
  theme(
    panel.border = element_blank(), 
    panel.grid.minor = element_blank()
  ) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  labs(
    y = "Fixed effect coefficient", 
    x = "", 
    title = "Reward signal sum"
  ) + 
  coord_flip()

ggsave(paste0(img_dir, "lme_sum.png"), width = 5, height = 2.5)
```

```{r}
ggplot(lme_avg_df, aes(x = coef_nm, y = fixfx)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi), width = 0) + 
  geom_errorbar(
    aes(ymin = fixfx - sd, ymax = fixfx + sd), 
    width = 0,
    size = 1
  ) + 
  theme(
    panel.border = element_blank(), 
    panel.grid.minor = element_blank()
  ) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  labs(
    y = "Fixed effect coefficient", 
    x = "", 
    title = "Reward signal average"
  ) + 
  coord_flip()

ggsave(paste0(img_dir, "lme_avg.png"), width = 5, height = 2.5)
```


